Statistical Grand Slams

### 2023-12-15 Lahman Analysis/Formal Analysis/Draft1.qmd
FINISHED

Furthermore, it is logical to ask: why not consider both? Incorporating more predictors can often make a model more accurate at prediction; and both Fellingham and Fisher (2017) and this paper are concerned with maximizing predictive power of our respective models. The reason this paper does not consider both is because of multicollinearity. Multicollinearity is where two independent variables in a model are highly or perfectly correlated and the model is unable to determine which variable is responsible for the corresponding change in the depending variable. Without a doubt there is a strong correlation between PA's and AB's because one is a subset of the other. In fact, we can see that the correlation between the two is quite strong in @fig-BattingCor. If we included both PA's and AB's in our model, our model's algorithm Markov Chain Monte Carlo (MCMC) would have a difficult time determining which of the two variables is responsible for the changes in the dependent variable, HR's. For other generalized linear models (GLM's) it becomes impossible to deal with multicollinearity and either the model parameters need to be changed or the model is deemed impossible to calculate. Thus, we neglect PA's from our model both because theory stipulates it is an inferior predictor and used in tandem with our current predictor it would cause problems for the modeling software.

@@

MLB has existed since about 1871. Since its inception the game has undergone some incredible transformations which have vastly changed the nature of the game, making it into what we know today. For example, from 1900-1920 baseball was in the "deadball era." During this period the materials used to create baseballs made it so that the ball absorbed much more force than it does today. This had a plethora of implications for the game and how it was played; for example, HR's were much less common then they are today because they were harder to hit. In contrast, the period from 1994-2004 is known informally as the "steroid era" because of the rampant use of performance enhancing drugs (PED) used by players. This, in combination with other factors such as harder materials used to make baseballs, led to players obliterating a variety of previous hitting records, such as the career and season home runs records. It is obvious from these facts that the decade played for player's had a drastic impact on their ability to produce some outcomes, such as HR's. In fact, this can be visualized by considering all the runs scored during a given season from 1871-2022 displayed in @fig-RunsByYear, or considering the total HR's hit from 1871-2022 displayed in @fig-HRByYear.

Finally, age is an important factor in athletic ability. In MLB, seniority can act both as both a blessing and a curse. On the one hand, older players are less able to hit home runs because their body is starting to lose its peak athletic performance compared to when it was, say, 30. On the other hand, these older players also have more experience than younger players and thus may be better able to hit home runs from strategies they have found during their time playing. Further we can visualize this by looking at the median HR's hit by each player from 1871-2022, which is displayed in @fig-HR.By.Age.All. However, this graph does not depict some of the nuances of the data well. For example, we have almost no data on how often 18 -year-old players hit HR's because they play in MLB infrequently. As can be seen with the standard error bars on the ages above 40, older players suffer a similar problem of small sample size. Instead, consider @fig-HR.By.Age.20to40 which only displays players who are 20-40 years old during the season. This graph clearly shows the complex non-linear relationship between player performance and age. It shows that players are generally improving in their ability to hit HR's from 20-30 years old. Then, however, we see a slight downwards trend followed by an upward trend with increasing error bars from ages 30-40. This shows that as players age past 30 their abilities start to decline. However, the jump afterwards can probably be attributed to players who are very good, continuing their careers. If a player is very good then they are going to be able to perform at above average when they are, say, 35. However, if the player is not generally considered good, then a team will likely opt to try out a new young rookie who may not be as good in the moment but is more projectable. Thus, we see less players playing past 35 which explains the higher standard errors, but they are of a higher caliber than any other age group which is why they are still playing. Thus, we see their proportion as higher than we would expect, given the trend of other 30-35 year old players.

@@

The parameter $\alpha_n$ (which is $\theta_i$ in the actual paper) are the individual intercepts for each player. This parameter is the only one that Fellingham and Fisher (2017) use which takes advantage of multilevel modeling. In this modeling, Fellingham and Fisher (2017) set the mean of $\alpha_n$ to be $\mu_d$ where $d$ is the decade the player was born. That means that Fellingham and Fisher (2017) set the parameter which measures the player's "innate" HR probability based on the HR probability of other players who were born in the same decade. This allows Fellingham and Fisher (2017) to make confident predictions about players who may only have 6 seasons of data because this model borrows data from all players who were born at around a similar time. Implicitly, this means that Fellingham and Fisher (2017) assume that when a player was born is a powerful predictor of their HR probability.

In contrast, the parameter $\xi_i$ (which is $\delta_i$ in the actual paper) represents the effect that season $i$ has on player $n$'s ability to hit a HR at stadium $p$. This $\xi$ is calculated for each season considered where $i\in\{1,…,138\}$ for the 138 seasons considered. This $\xi_i$ is calculated independent of each player so it represents the different changes that we see in HR ability over all years considered. An example of why this effect is important refer to @fig-HRByYear, which clearly shows how HR rates have differed drastically over the years considered. Although, the inclusion of $\xi_i$ is confusing because it probably captures the same effect that $\alpha_n$ captures. Most players in the MLB are within 10 years of age of each other with the most extreme different in age being about 20 years. Thus, the $\alpha_n$ of players is unlikely to differ by very much and any difference in $\alpha_n$ which appears is likely also apparent in $\xi_i$. Thus, the inclusion of both $xi_i$ and $\alpha_n$ is strange because it introduces a strong possibility of multicollinearity since both parameters describe roughly the same trend in the data.

@@

Another questionable choice that Fellingham and Fisher (2017) make is that they estimated their model parameters using FORTRAN. Given the antiquity of FORTRAN, replication of Fellingham and Fisher (2017) is quite difficult. Not to mention that the 200,000 iterations makes the research so computationally expensive that it is nearly impossible for any other research team to replicate. This is disappointing because the scientific method relies on replication for the advancement of knowledge. Thus, our research seeks to make the algorithms used in this research more accessible because we chose to use R to interface with the modeling software Stan, which is written in the (more contemporary) C++ programming language. Furthermore, our model will use less iterations and stronger priors so that other research teams can more easily replicate our findings.

@@

Exploring concretely how our model works will make it clearer than discussing it in the abstract. The current model that we have is the following:

@@

As discussed in @sec-BaseBack, there are many factors that influence player $n$'s underlying probability of hitting a $HR$ in a given $AB$. The first parameter which affects this probability, $\pi_{nip}$, is $\alpha_n$. In our model $\alpha_n$ can generally be thought of as the intercept term which is used to account for a player's innate ability to hit HR's. From our exploratory observations of the data, we noticed that the average (median or mean) proportion of AB's which resulted in HR's is about 0.03. This means that about 3% AB's result in HR's. This can be seen in @fig-HR.By.Age.20to40, @fig-pDist, and @fig-pDistSmall. On the logit scale, which is what our model relies on for estimation, a probability of 0.03 is about -3.5. We recognize that some players, for whatever reason, have better "innate" abilities to hit HR's than others. Therefore, we used a multilevel modeling technique for $\alpha_n$ to capture that player-to-player difference. To begin capturing this difference in $\alpha_n$ across players, we set the hyper-priors for $\alpha_n$ to be $\mu_0$ and $\sigma_0$. Where $\mu_0\sim Normal(-3.5,0.1)$ and $\sigma_0\sim Exponential(1)$. We set the mean of $\mu_0$ to be -3.5 because we estimate $\alpha_n$ on the logistic scale and we want it to be close to the average player's value. Since the parameters of $\alpha_n$ are distributions, we can think of $\alpha_n$ as a set of distributions, for the different values that either $\mu_0$ or $\sigma_0$ could take on. Further, the likelihood of the particular distribution that $\alpha_n$ takes on is dependent on the likelihood of either $\mu_0$ or $\sigma_0$ taking on a particular value. For example, $\mu_0=-3.0$ is extremely unlikely because it lies four standard deviations away from the mean of its normal distribution -3.5. Thus a distribution for $\alpha_n$ being $Normal(-3,\sigma_0)$ is highly unlikely for any value of $\sigma_0$. We chose $\mu_0$ to be normally distributed because there is precedent in Fellingham and Fisher (2017) to use normal distributions for their mean hyper-priors. Another reason it is easier to interpret normal distributions. We chose $\sigma_0$ to be exponentially distributed because the exponential distribution does not allow for negative values and the bulk of the possible values of $\sigma_0$ would be clustered near 0. Since we cannot have negative values for a standard deviation it seemed appropriate. This use of these hyper-priors is what makes the model appear to "learn". The model already comes in with certain expectations of what values $\alpha_n$ should take on, rather than simply coming into each new $n$ knowing nothing, which is what happens with more traditional GLM's. Finally, with this new set of distributions for $\alpha_n$ being the priors, our model constructs a posterior using Bayes theorem and the data on player $n$. It repeats this process for all 657 players that we considered for this research question to create a different $\alpha$ distribution for each player, $n$. Whatever value in $\alpha_n$ has the maximum likelihood, that's what we used to generate our predictions for inference in @sec-Results.

The next parameter we considered which affects $\pi_{nip}$ was $\beta_n$ which represents the effect centered age (the player's age minus 30) had on player $n$'s ability to hit HR's. This topic was briefly touched upon in @sec-BaseBack, where we discussed how a player's age can impact their their strategy and physical prowess. These ideas are well demonstrated in @fig-HR.By.Age.20to40. In @fig-HR.By.Age.20to40, one can easily see that, in general, players' ability to hit HR's improves over the course of their career and peaks when players reach about 30 years old. After this peak, a player's ability starts to decline for the rest of their career. We chose to center the the age of the all players at 30 for several reasons. The first was that roughly the mean and the median ages in the data was 30. Second, having a centered age makes interpreting $\beta_n$ easier. The value of $\beta_n$ represents how changes away from the centered age affects player $n$'s probability of hitting a HR, $\pi_{nip}$. In a normal GLM's, $\beta_n$ would not be able to adapt to each player, $n$, as it does in our model. It would only be able to consider how aging affects the average player's HR probability and then apply that effect to all players. This implicitly assumes that each player is affected identically by changes in age. In contrast, we employ multilevel modeling techniques to our model where $\beta_n\sim Normal(\mu_1,\sigma_1)$, $\mu_1\sim Normal(0,0.1)$, and $\sigma_1\sim Exponential(10)$. Similar to understanding $\alpha_n$, we can interpret $\beta_n$ as a collection of distributions corresponding to the different values of its hyper-priors $\mu_1$ and $\sigma_1$ and their corresponding likelihoods. We chose a normal distribution for $\mu_1$ for reasons similar to $\mu_0$. It makes interpretation easier and there is precedent in the literature. Furthermore, we chose it to have a mean of 0 because we assume that there is only a slightly positive increase in HR probability year over year, which is close enough to zero that we let the data speak for itself. Again, for similar reasons to $\sigma_0$ we chose $\sigma_1$ to be exponentially distributed. This ensured that $\sigma_1$ would not be negative and that the bulk of the possible values of $\sigma_1$ would be clustered near 0. With the distributions calculated for $\beta_n$ we can now use $\beta_n$ as the prior and combine it with the player data to generate the posterior using Bayes theorem. We repeat this process for all 657 players considered to get a unique distribution for each player $\beta_n$. This means that each player, $n$ has a unique aging coefficient which represents how aging effects their HR probability. Whatever value in $\beta_n$ has the maximum likelihood, that's what we used to generate our predictions for inference in @sec-Results.

>>>>

Furthermore, it is logical to ask: why not consider both? Incorporating more predictors can often make a model more accurate at prediction; and both Fellingham and Fisher (2017) and this paper are concerned with maximizing predictive power of our respective models. The reason this paper does not consider both is because of multicollinearity. Multicollinearity is where two independent variables in a model are highly or perfectly correlated and the model is unable to determine which variable is responsible for the corresponding change in the dependent variable. Without a doubt there is a strong correlation between PA's and AB's because one is a subset of the other. In fact, we can see that the correlation between the two is quite strong in @fig-BattingCor. If we included both PA's and AB's in our model, our model's algorithm Markov Chain Monte Carlo (MCMC) would have a difficult time determining which of the two variables is responsible for the changes in the dependent variable, HR's. For other generalized linear models (GLM's) it becomes impossible to deal with multicollinearity and either the model parameters need to be changed or the model is deemed impossible to calculate. Thus, we neglect PA's from our model both because theory stipulates it is an inferior predictor and used in tandem with our current predictor it would cause problems for the modeling software.

@@

MLB has existed since about 1871. Since its inception the game has undergone some incredible transformations which have vastly changed the nature of the game, making it into what we know today. For example, from 1900-1920 baseball was in the "deadball era." During this period the materials used to create baseballs made it so that the ball absorbed much more force than it does today. This had a plethora of implications for the game and how it was played; for example, HR's were much less common then they are today because they were harder to hit. In contrast, the period from 1994-2004 is known informally as the "steroid era" because of the rampant use of performance enhancing drugs (PED) used by players. This, in combination with other factors such as harder materials used to make baseballs, led to players obliterating a variety of previous hitting records, such as the career and season home runs records. It is obvious from these facts that the year played had a drastic impact on players' ability to produce some outcomes, such as HR's. In fact, this can be visualized by considering all the runs scored during a given season from 1871-2022 displayed in @fig-RunsByYear, or considering the total HR's hit from 1871-2022 displayed in @fig-HRByYear.

Finally, age is an important factor in athletic ability. In MLB, seniority can act both as both a blessing and a curse. On the one hand, older players are less able to hit HR's because their body is starting to lose its peak athletic performance compared to when it was, say, 30. On the other hand, these older players also have more experience than younger players and thus may be better able to hit HR's from strategies they have found during their time playing. Further we can visualize this by looking at the median HR's hit by each player from 1871-2022, which is displayed in @fig-HR.By.Age.All. However, this graph does not depict some of the nuances of the data well. For example, we have almost no data on how often 18 year old players hit HR's because they play in MLB infrequently. As can be seen with the standard error bars on the ages above 40, older players suffer a similar problem of small sample size. Instead, consider @fig-HR.By.Age.20to40 which only displays players who are 20-40 years old during the season. This graph clearly shows the complex non-linear relationship between player performance and age. It shows that players are generally improving in their ability to hit HR's from 20-30 years old. Then, however, we see a slight downwards trend followed by an upward trend with increasing error bars from ages 30-40. This shows that as players age past 30 their abilities start to decline. However, the jump afterwards can probably be attributed to players who are very good, continuing their careers. If a player is very good then they are going to be able to perform at above average when they are, say, 35. However, if the player is not generally considered good, then a team will likely opt to try out a new young rookie who may not be as good in the moment but is more projectable. Thus, we see less players playing past 35 which explains the higher standard errors, but they are of a higher caliber than any other age group which is why they are still playing. Thus, we see their proportion as higher than we would expect, given the trend of other 30-35 year old players.

@@

The parameter $\alpha_n$ (which is $\theta_i$ in the actual paper) are the individual intercepts for each player. This parameter is the only one that Fellingham and Fisher (2017) use which takes advantage of multilevel modeling. In this modeling, Fellingham and Fisher (2017) set the mean of $\alpha_n$ to be $\mu_d$ where $d$ is the decade the player was born. That means that Fellingham and Fisher (2017) set the parameter which measures the player's "innate" HR probability based on the HR probability of other players who were born in the same decade. This allows Fellingham and Fisher (2017) to make confident predictions about players who may only have 6 seasons of data because this model borrows data from all players who were born at around a similar time. Implicitly, this means that Fellingham and Fisher (2017) assume that when a player was born is a powerful predictor of their HR probability, which seems questionable.

In contrast, the parameter $\xi_i$ (which is $\delta_i$ in the actual paper) represents the effect that season $i$ has on player $n$'s ability to hit a HR in year $i$. This $\xi$ is calculated for each season considered where $i\in\{1,…,138\}$ for the 138 seasons considered. This $\xi_i$ is calculated independent of each player so it represents the different changes that we see in HR ability over all years considered. An example of why this effect is important refer to @fig-HRByYear, which clearly shows how HR rates have differed drastically over the years considered. Although, the inclusion of $\xi_i$ is confusing because it probably captures the same effect that $\alpha_n$ captures. Most players in the MLB are within 10 years of age of each other with the most extreme different in age being about 20 years. Thus, the $\alpha_n$ of players is unlikely to differ by very much and any difference in $\alpha_n$ which appears is likely also apparent in $\xi_i$. Thus, the inclusion of both $\xi_i$ and $\alpha_n$ is strange because it introduces a strong possibility of multicollinearity since both parameters describe roughly the same trend in the data.

@@

Another questionable choice that Fellingham and Fisher (2017) make is that they estimated their model parameters using FORTRAN. Given the antiquity of FORTRAN, replication of Fellingham and Fisher (2017) is quite difficult. Not to mention that the 200,000 iterations makes the research so computationally expensive that it is nearly impossible for any other research team to replicate. This is disappointing because the scientific method relies on replication for the advancement of knowledge. Thus, our research seeks to make an algorithm that is more accessible because we chose to use R to interface with the modeling software Stan, which is written in the (more contemporary) C++ programming language. Furthermore, our model will use less iterations and stronger priors so that other research teams can more easily replicate our findings.

@@

Our model is the following:

@@

As discussed in @sec-BaseBack, there are many factors that influence player $n$'s underlying probability of hitting a $HR$ in a given $AB$. The first parameter which affects this probability, $\pi_{nip}$, is $\alpha_n$. In our model $\alpha_n$ can generally be thought of as the intercept term which is used to account for a player's innate ability to hit HR's. From our exploratory observations of the data, we noticed that the average (median or mean) proportion of AB's which resulted in HR's is about 0.03. This means that about 3% AB's result in HR's. This can be seen in @fig-HR.By.Age.20to40, @fig-pDist, and @fig-pDistSmall. On the logit scale, which is what our model relies on for estimation, a probability of 0.03 is about -3.5. We recognize that some players, for whatever reason, have better "innate" abilities to hit HR's than others. Therefore, we used a multilevel modeling technique for $\alpha_n$ to capture that player-to-player difference. To begin capturing this difference in $\alpha_n$ across players, we set the hyper-priors for $\alpha_n$ to be $\mu_0$ and $\sigma_0$. Where $\mu_0\sim Normal(-3.5,0.1)$ and $\sigma_0\sim Exponential(1)$. We set the mean of $\mu_0$ to be -3.5 because we estimate $\alpha_n$ on the logistic scale and we want it to be close to the average player's value. Since the parameters of $\alpha_n$ are distributions, we can think of $\alpha_n$ as a set of distributions, for the different values that either $\mu_0$ or $\sigma_0$ could take on. Further, the likelihood of the particular distribution that $\alpha_n$ takes on is dependent on the likelihood of either $\mu_0$ or $\sigma_0$ taking on a particular value. For example, $\mu_0=-3.0$ is extremely unlikely because it lies five standard deviations away from the mean of its normal distribution -3.5. Thus a distribution for $\alpha_n$ being $Normal(-3,\sigma_0)$ is highly unlikely for any value of $\sigma_0$. We chose $\mu_0$ to be normally distributed because there is precedent in Fellingham and Fisher (2017) to use normal distributions for their mean hyper-priors. Another reason it is easier to interpret normal distributions. We chose $\sigma_0$ to be exponentially distributed because the exponential distribution does not allow for negative values and the bulk of the possible values of $\sigma_0$ would be clustered near 0. Since we cannot have negative values for a standard deviation it seemed appropriate. This use of these hyper-priors is what makes the model appear to "learn". The model already comes in with certain expectations of what values $\alpha_n$ should take on, rather than simply coming into each new $n$ knowing nothing, which is what happens with more traditional GLM's. Finally, with this new set of distributions for $\alpha_n$ being the priors, our model constructs a posterior using Bayes theorem and the data on player $n$. It repeats this process for all 657 players that we considered for this research question to create a different $\alpha$ distribution for each player, $n$. Whatever value in $\alpha_n$ has the maximum likelihood, that's what we used to generate our predictions for inference in @sec-Results.

The next parameter we considered which affects $\pi_{nip}$ was $\beta_n$ which represents the effect centered age (the player's age minus 30) had on player $n$'s ability to hit HR's. This topic was briefly touched upon in @sec-BaseBack, where we discussed how a player's age can impact their their strategy and physical prowess. These ideas are well demonstrated in @fig-HR.By.Age.20to40. In @fig-HR.By.Age.20to40, one can easily see that, in general, players' ability to hit HR's improves over the course of their career and peaks when players reach about 30 years old. After this peak, a player's ability starts to decline for the rest of their career. We chose to center the the age of the all players at 30 for several reasons. The first was that roughly the mean and the median ages in the data was 30. Second, having a centered age makes interpreting $\beta_n$ easier. The value of $\beta_n$ represents how changes away from the centered age affects player $n$'s probability of hitting a HR, $\pi_{nip}$. In normal GLM's, $\beta_n$ would not be able to adapt to each player, $n$, as it does in our model. It would only be able to consider how aging affects the average player's HR probability and then apply that effect to all players. This implicitly assumes that each player is affected identically by changes in age. In contrast, we employ multilevel modeling techniques to our model where $\beta_n\sim Normal(\mu_1,\sigma_1)$, $\mu_1\sim Normal(0,0.1)$, and $\sigma_1\sim Exponential(10)$. Similar to understanding $\alpha_n$, we can interpret $\beta_n$ as a collection of distributions corresponding to the different values of its hyper-priors $\mu_1$ and $\sigma_1$ and their corresponding likelihoods. We chose a normal distribution for $\mu_1$ for reasons similar to $\mu_0$. It makes interpretation easier and there is precedent in the literature. Furthermore, we chose it to have a mean of 0 because we assume that there is only a slightly positive increase in HR probability year over year, which is close enough to zero that we let the data speak for itself. Again, for similar reasons to $\sigma_0$ we chose $\sigma_1$ to be exponentially distributed. This ensured that $\sigma_1$ would not be negative and that the bulk of the possible values of $\sigma_1$ would be clustered near 0. With the distributions calculated for $\beta_n$ we can now use $\beta_n$ as the prior and combine it with the player data to generate the posterior using Bayes theorem. We repeat this process for all 657 players considered to get a unique distribution for each player $\beta_n$. This means that each player, $n$ has a unique aging coefficient which represents how aging effects their HR probability. Whatever value in $\beta_n$ has the maximum likelihood, that's what we used to generate our predictions for inference in @sec-Results.
